Assessing the Firing of Ceramic Materials: a Seriation-Based Approach

### 2023-08-01 analysis/paper/paper.qmd
Edits after reviews.

---
title: "Assessing the Firing of Ceramic Materials: a Seriation-Based Approach"
author:
  - Nicolas Frerebeau:
      correspondence: "yes"
      email: nicolas.frerebeau@u-bordeaux-montaigne.fr
      orcid: 0000-0001-5759-4944
      affiliations:
        - name: "UMR 6034 Archéosciences Bordeaux"
          address: "Maison de l'Archéologie, Université Bordeaux Montaigne"
          city: Pessac cedex
          country: France
          postal-code: 33607
          url: https://www.archeosciences-bordeaux.fr
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  pdf: 
    documentclass: scrartcl
    include-in-header: 
      - text: |
          \usepackage{booktabs}
    classoption: 
      - french
      - 12pt
      - a4paper
      - oneside
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    comment: "#>"
    fig.path: "../figures/"
    fig.dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: [references.bib, packages.bib]
csl: "../templates/archeosciences.csl"
abstract: |
  The study of the firing of archaeological ceramics is a key factor in 
  interpreting ancient techniques and related social practices. This study 
  proposes a new method based on the correspondence analysis of powder X-ray 
  diffraction data (XRD), aiming to provide an alternative to the so-called
  Equivalent Firing Temperature framework. The properties of correspondence 
  analysis and its application for matrix seriation enable the description of 
  the firing of a ceramic assemblage. The application of correspondence 
  analysis to XRD data of Iron Age ceramics revealed the  significant potential 
  of this statistical method in the technological analysis of ancient pottery. 
  This allows to order of a large number of samples according to a firing 
  gradient and to track mineralogical transformations occurring during firing.
  This method should let to more precise comparisons than those based on 
  estimating the equivalent firing temperature.
keywords: |
  X-ray diffraction; ceramic firing; correspondence analysis; matrix seriation
---

@@

| **Version** | **Date**   | **Description**            | **DOI** |
|------------:|-----------:|:---------------------------|:--------|
| 0.0.0.9000  |  | Submitted to _Archéosciences_ |  |

@@

Correspondence analysis (CA) is similar to principal component analysis (PCA) in that it allows describing the statistical relationships that can exist between individuals and variables. However, the concept of similarity between rows and columns is different. In CA, two rows (resp. columns) are close to each other if they associate with columns (resp. rows) in the same way. Unlike PCA, CA analyzes the differences between relative values and considers both individuals and variables at the same time (this allows for the projection of row and column points in the same coordinate space). CA studies the inertia (i.e., the weighted sum of the χ2 distances of each point to the centroid) to create orthogonal components so that a maximum of the total inertia is represented on the first component, a maximum of the residual inertia on the second component, and so on until the last dimension. CA maximizes the correspondence between individuals and variables instead of maximizing the amount of variance explained by its reduced space.

CA is an effective method for the chronological seriation of archaeological assemblages [see @ihm2005 for an historical overview]. The order of the rows and columns is given by the coordinates along one dimension of the CA space, assumed to account for temporal variation. The direction of temporal change within the correspondence analysis space is arbitrary: additional information is needed to determine the actual order in time. CA is not limited to chronological modeling but is widely used to find a possible arrangement (ordering) of individuals and variables along any gradient, i.e., any aspect that is expected to be related to the observed composition (be it temporal or environmental).

@@

The results of the correspondence analysis of the diffractograms are presented in @fig-plot-ca. Only the data from the analysis of the samples from Mas de Moreno were used, including unfired ceramics and shards. The data from El Palao and Torre Cremada were introduced into the analysis as supplementary individuals. These additional individuals do not contribute to the calculation of the correspondence analysis itself, but they are projected onto the existing analysis space to visualize their relationships with the original dataset.

@@

## Columns (F1 vs F2)
# plot(
#   x = xrd_moreno_col_coord[, 1],
#   y = xrd_moreno_col_coord[, 2],
#   xlim = xlim, ylim = ylim,
#   xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
#   ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
#   type = "p",
#   pch = 16,
#   las = 1, asp = 1,
#   col = adjustcolor("black", alpha = 0.2)
# )
# text(x = min(xlim), y = max(ylim) + 0.25, labels = "D", font = 2, cex = 1.5)
# abline(h = 0, lty = 2, col = "darkgrey")
# abline(v = 0, lty = 2, col = "darkgrey")
col_minerals <- khroma::color("bright")(6)
plot(
  x = xrd_minerals_coord[, 1],
  y = xrd_minerals_coord[, 2],
  xlim = xlim, ylim = ylim,
  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
  type = "p",
  pch = 16,
  las = 1, asp = 1,
  col = col_minerals[as.factor(minerals$symbole)]
)
abline(h = 0, lty = 2, col = "darkgrey")
abline(v = 0, lty = 2, col = "darkgrey")
text(
  x = xrd_minerals_coord[, 1],
  y = xrd_minerals_coord[, 2],
  labels = mapply(
    FUN = function(s, h, k, l) {
      if (h < 0) as.expression(bquote(.(s)~"("*bar(.(abs(h)))*.(k)*.(l)*")"))
      else as.expression(bquote(.(s)~"("*.(h)*.(k)*.(l)*")"))
    }, 
    s = minerals$symbole,
    h = minerals$h, 
    k = minerals$k,
    l = minerals$l
  ),
  col = col_minerals[as.factor(minerals$symbole)],
  pos = c(2,3, 2, 1, 4, 4, 4, 2, 2, 2, 4, 3, 3, 4, 3, 1, 4)
)
text(x = min(xlim), y = max(ylim) + 0.25, labels = "B", font = 2, cex = 1.5)
```

@@

The first plane of the analysis exhibits the characteristic Guttman effect, also known as the arch effect (@fig-plot-ca, @fig-plot-guttman). This effect is characterized by a strong nonlinear relationship between the second factor and the first factor. The Guttman effect typically arises when a single dominant latent variable is present [@lebart2014]. Interestingly, the samples located in the upper right quadrant of the correspondence analysis plot are linked to diffraction peaks that correspond to clay minerals and calcite (@fig-plot-guttman A, @fig-plot-ca D, @fig-plot-xrd). These samples are positioned opposite to the ones associated with the peaks of diopside and anorthite along axis 1. Conversely, the samples located in the two lower quadrants are linked to the peaks of quartz and the melilite series.

@@

The material from the Mas de Moreno is the result of waste from the production process. It is therefore necessary to distinguish between what is representative of material produced in the workshop and what is the result of a non-desirable situation leading to deliberate rejection (firing defect). In order to explore these two hypotheses, the diffratograms of several artifacts from consumption contexts (El Palao and Torre Cremada) were introduced into the previous analysis as supplementary individuals. This was made possible by the technical homogeneity of the material studied. Samples from El Palao and Torre Cremada have the same compositional ranges (@fig-plot-guttman D) and belong to the same category of Iberian fine wares as the material from Mas de Moreno.

>>>>

---
title: "Assessing the Firing of Ceramic Materials: a Seriation-Based Approach"
author:
  - Nicolas Frerebeau:
      correspondence: "yes"
      email: nicolas.frerebeau@u-bordeaux-montaigne.fr
      orcid: 0000-0001-5759-4944
      affiliations:
        - name: "UMR 6034 Archéosciences Bordeaux"
          address: "Maison de l'Archéologie, Université Bordeaux Montaigne"
          city: Pessac cedex
          country: France
          postal-code: 33607
          url: https://www.archeosciences-bordeaux.fr
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  pdf: 
    documentclass: scrartcl
    include-in-header: 
      - text: |
          \usepackage{booktabs}
    classoption: 
      - french
      - 12pt
      - a4paper
      - oneside
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
knitr:
  opts_chunk:
    echo: false
    warning: false
    message: false
    comment: "#>"
    fig.path: "../figures/"
    fig.dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: [references.bib, packages.bib]
csl: "../templates/archeosciences.csl"
abstract: |
  The study of the firing of archaeological ceramics is a key factor in 
  interpreting ancient techniques and related social practices. This study 
  proposes a new method based on the correspondence analysis of powder X-ray 
  diffraction data (XRD), aiming to provide an alternative to the so-called
  Equivalent Firing Temperature framework. The properties of correspondence 
  analysis and its application for matrix seriation enable the description of 
  the firing of a ceramic assemblage. The application of correspondence 
  analysis to XRD data of Iron Age ceramics revealed the  significant potential 
  of this statistical method in the technological analysis of ancient pottery. 
  This allows to order a large number of samples according to a firing 
  gradient and to track mineralogical transformations occurring during firing.
  This method should let to more precise comparisons than those based on 
  estimating the equivalent firing temperature.
keywords: |
  X-ray diffraction; ceramic firing; correspondence analysis; matrix seriation
---

@@

| **Version** | **Date**   | **Description**            | **DOI** |
|---------:|-----------:|:------------------------------|:---------------------------|
| 1.0      | 2023-06-09 | Submitted to _Archéosciences_ | [10.5281/zenodo.8020137](https://doi.org/10.5281/zenodo.8020137) |
| 1.1      | 2023-08-01 | Corrections based on reviews  | |

@@

Correspondence analysis (CA) is similar to principal component analysis (PCA) in that it allows describing the statistical relationships that can exist between individuals and variables. However, the concept of similarity between rows and columns is different. In CA, two rows (resp. columns) are close to each other if they associate with columns (resp. rows) in the same way. Unlike PCA, CA analyzes the differences between relative values and considers both individuals and variables at the same time. This allows for the projection of row and column points in the same coordinate space, so that the relative positions of one set of points provide the reading keys for interpreting the positions of the other set of points along the axes [@greenacre2007].

CA studies the inertia (i.e., the weighted sum of the $\chi^2$ distances of each point to the centroid) to create orthogonal components so that a maximum of the total inertia is represented on the first component, a maximum of the residual inertia on the second component, and so on until the last dimension. CA maximizes the correspondence between individuals and variables instead of maximizing the amount of variance explained by its reduced space.

CA is an effective method for the chronological seriation of archaeological assemblages [see @ihm2005 for an historical overview]. The order of the rows and columns is given by the coordinates along one dimension of the CA space, assumed to account for temporal variation. The direction of temporal change within the correspondence analysis space is arbitrary: additional information is needed to determine the actual order in time. CA is not limited to chronological modelling but is widely used to find a possible arrangement (ordering) of individuals and variables along any gradient, i.e., any aspect that is expected to be related to the observed composition (be it temporal or environmental).

@@

The results of the correspondence analysis of the diffractograms are presented in @fig-plot-ca. Only the data from the analysis of the samples from Mas de Moreno were used, including unfired ceramics and sherds. The data from El Palao and Torre Cremada were introduced into the analysis as supplementary individuals. These additional individuals do not contribute to the calculation of the correspondence analysis itself, but they are projected onto the existing analysis space to visualize their relationships with the original dataset.

@@

## Columns (F1 vs F2)
# plot(
#   x = xrd_moreno_col_coord[, 1],
#   y = xrd_moreno_col_coord[, 2],
#   xlim = xlim, ylim = ylim,
#   xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
#   ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
#   type = "p",
#   pch = 16,
#   las = 1, asp = 1,
#   col = adjustcolor("black", alpha = 0.2)
# )
# text(x = min(xlim), y = max(ylim) + 0.25, labels = "D", font = 2, cex = 1.5)
# abline(h = 0, lty = 2, col = "darkgrey")
# abline(v = 0, lty = 2, col = "darkgrey")
col_minerals <- khroma::color("bright")(6)
plot(
  x = xrd_minerals_coord[, 1],
  y = xrd_minerals_coord[, 2],
  xlim = xlim, ylim = ylim,
  xlab = sprintf("F1 (%g%%)", round(xrd_moreno_eig[1, 2], 1)),
  ylab = sprintf("F2 (%g%%)", round(xrd_moreno_eig[2, 2], 1)),
  type = "p",
  pch = 16,
  las = 1, asp = 1,
  col = col_minerals[as.factor(minerals$symbole)]
)
abline(h = 0, lty = 2, col = "darkgrey")
abline(v = 0, lty = 2, col = "darkgrey")
text(
  x = xrd_minerals_coord[, 1],
  y = xrd_minerals_coord[, 2],
  labels = mapply(
    FUN = function(s, h, k, l) {
      if (h < 0) as.expression(bquote(.(s)~"("*bar(.(abs(h)))*.(k)*.(l)*")"))
      else as.expression(bquote(.(s)~"("*.(h)*.(k)*.(l)*")"))
    }, 
    s = minerals$symbole,
    h = minerals$h, 
    k = minerals$k,
    l = minerals$l
  ),
  col = col_minerals[as.factor(minerals$symbole)],
  pos = c(2,3, 2, 1, 4, 4, 4, 2, 2, 2, 4, 3, 3, 4, 3, 1, 4)
)
text(x = min(xlim), y = max(ylim) + 0.25, labels = "D", font = 2, cex = 1.5)
```

@@

The first plane of the analysis exhibits the characteristic Guttman effect, also known as the arch effect (@fig-plot-ca, @fig-plot-guttman). This effect is characterized by a strong nonlinear relationship between the second factor and the first factor. The Guttman effect typically arises when a single dominant latent variable is present [@lebart2014]. Interestingly, the samples located in the upper left quadrant of the correspondence analysis plot are linked to diffraction peaks that correspond to clay minerals and calcite (@fig-plot-guttman A, @fig-plot-ca D, @fig-plot-xrd). These samples are positioned opposite to the ones associated with the peaks of diopside and anorthite along axis 1. Conversely, the samples located in the two lower quadrants are linked to the peaks of quartz and the melilite series.

@@

The material from the Mas de Moreno is the result of waste from the production process. It is therefore necessary to distinguish between what is representative of material produced in the workshop and what is the result of a non-desirable situation leading to deliberate rejection (firing defect). In order to explore these two hypotheses, the diffractograms of several artifacts from consumption contexts (El Palao and Torre Cremada) were introduced into the previous analysis as supplementary individuals. This was made possible by the technical homogeneity of the material studied. Samples from El Palao and Torre Cremada have the same compositional ranges (@fig-plot-guttman D) and belong to the same category of Iberian fine wares as the material from Mas de Moreno.
